Human Factors Influencing Visual Statistical Inference

### 2013-06-06 sociology_paper.tex
revised abstract

\begin {abstract}  
Visual Statistical Inference is dependent on the careful evaluation of the lineups by individual observers. Each individual is different in their cognitive phycology and judiciousness which can affect the power of visual inference. To estimate the power of visual inference this can be controlled by getting evaluations from multiple people from a diverse population. But the other factor that may affect the power of visual inference is the demographics and skills of the observer. In this paper we examine this in details. The simulation experiments suggest that individual skill as well as demographics are very significant for the power of visual inference. We examine the evidence of learning or getting skilled while sequentially evaluating multiple lineups which enables them evaluate a lineup faster in the later attempts. The experimental data suggest that there is no learning effect in terms of correct response. The location of actual plot in the lineups does not have any significant effect as well.

{\bf Keywords: \sf statistical graphics, lineup, non-parametric test, cognitive phycology, visualization, exploratory data analysis} 
\end {abstract}

>>>>

\begin {abstract}  
Visual statistical inference is a way to determine significance of patterns found while exploring data. It is dependent on the evaluation of a lineup, of a data plot among a sample of null plots, by human observers. Each individual is different in their cognitive psychology and judiciousness, which can affect the visual inference. The usual way to estimate the effectiveness of a statistical test is its power. The estimate of power of a lineup can be controlled by combining evaluations from multiple observers. Factors that may also affect the power of visual inference are the observers' demographics, visual skills, and experience, the sample of null plots taken from the null distribution, the position of the data plot in the lineup, and the signal strength in the data. This paper examine these factors. Results from multiple visual inference studies are examined to provide an assessment of these. The experiments suggest that individual skills vary substantially,  but demographics do not have a huge effect. There is evidence that a learning effect exists that observers get better and faster with repeated evaluations which suggests a potential for using the lineup protocol for training people to read statistical graphics. The placement of actual data plot in the lineups does not affect the inference.

{\bf Keywords: \sf statistical graphics, non-parametric test, cognitive phycology, data visualization, exploratory data analysis, data mining, visual analytics} 
\end {abstract}
